Best practices for forecasting changes in biodiversity: an example using breeding birds

### 2017-09-05 manuscript/draft.Rmd

In addition to comparing and improving the process models used for forecasting
it is important to consider the observation models [Box 1].  When working with
any ecological dataset, there are imperfections in the sampling process that
have the potential to influence results. With large scale surveys and citizen
science datasets, such as the Breeding Bird Survey, these issues are potentially
magnified by the large number of different observers and by major differences in
the habitats and species being surveyed [@sauer1994]. Accounting for differences
in observers reduced the average error in our point estimates and also improved
the coverage of the confidence intervals. In addition, controlling for observer
effects resulted in changes in which models performed best, most notably
improving most models' point estimates relative to the naive baseline. This
suggests that, prior to accounting for observer effects, the naive model
performed well largely because it was capable of accommodating rapid shifts in
estimated richness introduced by changes in the observer. These kinds of rapid
changes were difficult for the other single-site models to accommodate. Another
key aspect of an ideal observation model is imperfect detection. In this study,
we did not address differences in detection probability across species and sites
[@boulinier1998] since there is no clear way to address this issue using North
American Breeding Bird Survey data without making strong assumptions about the
data [i.e., assuming there is no biological variation in stops along a route;
@white2010], but this would be a valuable addition to future forecasting models.

>>>>

In addition to comparing and improving the process models used for forecasting
it is important to consider the observation models.  When working with any
ecological dataset, there are imperfections in the sampling process that have
the potential to influence results. With large scale surveys and citizen science
datasets, such as the Breeding Bird Survey, these issues are potentially
magnified by the large number of different observers and by major differences in
the habitats and species being surveyed [@sauer1994]. Accounting for differences
in observers reduced the average error in our point estimates and also improved
the coverage of the confidence intervals. In addition, controlling for observer
effects resulted in changes in which models performed best, most notably
improving most models' point estimates relative to the naive baseline. This
demonstrates that modeling observation error can be important for properly
estimating and reducing uncertainty in forecasts and can also lead to changes in
the best methods for forecasting [Box 1]. This suggests that, prior to
accounting for observer effects, the naive model performed well largely because
it was capable of accommodating rapid shifts in estimated richness introduced by
changes in the observer. These kinds of rapid changes were difficult for the
other single-site models to accommodate. Another key aspect of an ideal
observation model is imperfect detection. In this study, we did not address
differences in detection probability across species and sites [@boulinier1998]
since there is no clear way to address this issue using North American Breeding
Bird Survey data without making strong assumptions about the data [i.e.,
assuming there is no biological variation in stops along a route; @white2010],
but this would be a valuable addition to future forecasting models.
